---
title: "Getting started with ggsunburstR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ggsunburstR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6
)
```

## Overview

ggsunburstR creates sunburst and icicle plots from hierarchical data using
ggplot2. It accepts multiple input formats and produces standard ggplot2
objects you can customise freely.

## Basic usage

The workflow has two steps:

1. **Prepare data** with `sunburst_data()`
2. **Plot** with `sunburst()`, `icicle()`, `donut()`, or `ggtree()`

```{r basic}
library(ggsunburstR)

# Parse a Newick tree string
sb <- sunburst_data("((a, b, c), (d, e, f));")

# Sunburst plot
sunburst(sb, fill = "depth")
```

```{r icicle}
# Same data as an icicle plot
icicle(sb, fill = "depth")
```

## Input formats

### Newick strings

The most common input for phylogenetic or hierarchical trees:

```{r newick}
sb <- sunburst_data("((mammals, birds, reptiles)vertebrates, (beetles, flies)insects)animals;")
sunburst(sb, fill = "name")
```

### Data frames

Use a data frame with `parent` and `child` columns:

```{r dataframe}
df <- data.frame(
  parent = c(NA, "root", "root", "fruits", "fruits", "vegs", "vegs"),
  child  = c("root", "fruits", "vegs", "apple", "banana", "carrot", "pea")
)
sb <- sunburst_data(df)
sunburst(sb, fill = "name")
```

### Path-delimited strings

Hierarchies can also be expressed as slash-delimited paths:

```{r paths}
sb <- sunburst_data(c("A/B/C", "A/B/D", "A/E/F"))
sunburst(sb, fill = "name")
```

### Other formats

ggsunburstR also accepts Newick files, node-parent CSVs, lineage TSVs,
`ape::phylo` objects, and `data.tree::Node` objects. See `?sunburst_data`
for details.

## Plot types

### Donut charts

A donut is a sunburst restricted to the outermost depth levels:

```{r donut}
sb <- sunburst_data("((a, b, c), (d, e, f));")
donut(sb, fill = "name", levels = 1, hole_size = 2)
```

### Tree dendrograms

`ggtree()` renders the tree as a dendrogram:
```{r ggtree}
sb <- sunburst_data("((a:0.3, b:0.5)X:0.2, (c:0.4, d:0.1)Y:0.3)root;")
ggtree(sb, show_labels = TRUE)
```

## Value-weighted sectors

By default, all leaves get equal angular width. Use `values` to size
sectors proportionally:

```{r values}
sb <- sunburst_data(
  "((a, b, c), (d, e));",
  values = c(a = 10, b = 5, c = 3, d = 8, e = 2)
)
sunburst(sb, fill = "name")
```

## Fill mapping

The `fill` parameter accepts column names (quoted or bare), plus the
special values `"auto"` (maps to depth) and `"none"` (static grey):

```{r fill-auto}
sb <- sunburst_data("((a, b, c), (d, e, f));")

# Bare name (tidy eval)
sunburst(sb, fill = depth)
```

```{r fill-name}
# Auto-map to depth
sunburst(sb, fill = "auto")
```

## Labels

### Basic labels

Add leaf labels with `show_labels = TRUE`:

```{r labels}
sb <- sunburst_data("((a, b, c), (d, e, f));")
sunburst(sb, fill = "depth", show_labels = TRUE)
```

### Perpendicular labels

Use `label_type = "perpendicular"` for arc-following labels:

```{r perp-labels}
sunburst(sb, fill = "depth", show_labels = TRUE,
         label_type = "perpendicular")
```

### Internal node labels and filtering

Show labels for internal nodes and filter out labels on narrow sectors:

```{r node-labels}
sb <- sunburst_data("((mammals, birds, reptiles)vertebrates, (beetles, flies)insects)animals;")
sunburst(sb, fill = "name", show_labels = TRUE,
         show_node_labels = TRUE, min_label_angle = 30,
         label_size = 3.5)
```

### Label repulsion (icicle)

Use `ggrepel` for collision avoidance in icicle plots:

```{r repel-labels, eval = requireNamespace("ggrepel", quietly = TRUE)}
sb <- sunburst_data("((a, b, c, d, e), (f, g, h));")
icicle(sb, fill = "depth", show_labels = TRUE, label_repel = TRUE)
```

## Drilldown into subtrees

Zoom into a specific branch with `drilldown()`:

```{r drilldown}
sb <- sunburst_data("((a, b)X, (c, d)Y)root;")
sub <- drilldown(sb, node = "X")
sunburst(sub, fill = "name")
```

## Annotations

### Bar chart annotations

Add bar charts adjacent to leaf nodes:

```{r bars}
df <- data.frame(
  parent = c(NA, "root", "root", "root"),
  child  = c("root", "A", "B", "C"),
  score  = c(NA, 0.5, 0.9, 0.3)
)
sb <- sunburst_data(df)
p <- icicle(sb, fill = "name")
bars(p, sb, variables = "score")
```

### Tile (heatmap) annotations

Add colour-coded tiles adjacent to leaf nodes:

```{r tile}
df <- data.frame(
  parent = c(NA, "root", "root", "root"),
  child  = c("root", "A", "B", "C"),
  group  = c(NA, "x", "y", "x")
)
sb <- sunburst_data(df)
p <- icicle(sb, fill = "name")
tile(p, sb, variables = "group")
```

## Highlight specific nodes

Emphasise nodes by name or ID:

```{r highlight}
sb <- sunburst_data("((a, b, c), (d, e));")
p <- sunburst(sb, fill = "depth")
highlight_nodes(p, nodes = c("a", "c"), fill = "red")
```

## Per-depth fill scales

Use `sunburst_multifill()` or `icicle_multifill()` to map different
colour scales to different hierarchy levels:

```{r multifill, eval = requireNamespace("ggnewscale", quietly = TRUE)}
sb <- sunburst_data("((a, b)X, (c, d)Y)root;")
sunburst_multifill(sb, fills = list("1" = "name", "2" = "name"))
```

## ggplot2 geom

For idiomatic ggplot2 syntax, use `geom_sunburst()` directly with a
data.frame:

```{r geom-sunburst}
df <- data.frame(
  parent = c(NA, "root", "root", "A", "A"),
  child  = c("root", "A", "B", "a1", "a2"),
  group  = c("r", "g1", "g2", "g1", "g1")
)

ggplot2::ggplot(df) +
  geom_sunburst(ggplot2::aes(id = child, parent = parent, fill = group)) +
  ggplot2::coord_polar() +
  theme_sunburst()
```

## Tree inspection

Use `nw_print()` to inspect tree structure:

```{r nw-print}
nw_print("((a, b)X, (c, d)Y)root;")
```

## Customisation with ggplot2

Since the output is a standard ggplot2 object, you can add any ggplot2
layer or theme:

```{r custom}
sb <- sunburst_data("((a, b, c), (d, e, f));")
sunburst(sb, fill = "name") +
  ggplot2::scale_fill_brewer(palette = "Set2") +
  ggplot2::labs(title = "Custom sunburst") +
  theme_sunburst()
```

## Options

Key parameters for `sunburst_data()`:

- `values` — named numeric vector or column name for sector sizing
- `branchvalues` — `"remainder"` (default) or `"total"`
- `leaf_mode` — `"actual"` (default) or `"extended"`
- `ladderize` — sort by descendant count
- `ultrametric` — equalise leaf depths
- `xlim` — angular span (default 360)
- `rot` — rotation offset in degrees
