---
title: "Getting started with ggsunburstR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ggsunburstR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6
)
```

## Overview

ggsunburstR creates sunburst and icicle plots from hierarchical data using
ggplot2. It accepts multiple input formats and produces standard ggplot2
objects you can customise freely.

## Basic usage

The workflow has two steps:

1. **Prepare data** with `sunburst_data()`
2. **Plot** with `sunburst()` or `icicle()`

```{r basic}
library(ggsunburstR)

# Parse a Newick tree string
sb <- sunburst_data("((a, b, c), (d, e, f));")

# Sunburst plot
sunburst(sb, fill = "depth")
```

```{r icicle}
# Same data as an icicle plot
icicle(sb, fill = "depth")
```

## Input formats

### Newick strings

The most common input for phylogenetic or hierarchical trees:

```{r newick}
sb <- sunburst_data("((mammals, birds, reptiles)vertebrates, (beetles, flies)insects)animals;")
sunburst(sb, fill = "name")
```

### Data frames

Use a data frame with `parent` and `child` columns:

```{r dataframe}
df <- data.frame(
  parent = c(NA, "root", "root", "fruits", "fruits", "vegs", "vegs"),
  child  = c("root", "fruits", "vegs", "apple", "banana", "carrot", "pea")
)
sb <- sunburst_data(df)
sunburst(sb, fill = "name")
```

### Files

ggsunburstR can read Newick files, node-parent CSVs, and lineage TSVs
directly by passing a file path.

## Value-weighted sectors

By default, all leaves get equal angular width. Use `values` to size
sectors proportionally:

```{r values}
sb <- sunburst_data(
  "((a, b, c), (d, e));",
  values = c(a = 10, b = 5, c = 3, d = 8, e = 2)
)
sunburst(sb, fill = "name")
```

## Customisation with ggplot2

Since the output is a standard ggplot2 object, you can add any ggplot2
layer or theme:

```{r custom}
sb <- sunburst_data("((a, b, c), (d, e, f));")
sunburst(sb, fill = "name") +
  ggplot2::scale_fill_brewer(palette = "Set2") +
  ggplot2::labs(title = "Custom sunburst") +
  ggplot2::theme(legend.position = "bottom")
```

## Labels

Add leaf labels with `show_labels = TRUE`:

```{r labels}
sb <- sunburst_data("((a, b, c), (d, e, f));")
sunburst(sb, fill = "depth", show_labels = TRUE)
```

## Options

Key parameters for `sunburst_data()`:

- `values` — named numeric vector or column name for sector sizing
- `branchvalues` — `"remainder"` (default) or `"total"`
- `leaf_mode` — `"actual"` (default) or `"extended"`
- `ladderize` — sort by descendant count
- `ultrametric` — equalise leaf depths
- `xlim` — angular span (default 360)
- `rot` — rotation offset in degrees
